---
title: "Project 4 Code Base Submission"
author: "Long Lin, Zihao Yu"
format: html
editor: visual
---

## Overview

For this project, We used the documents within the ham/spam dataset located at https://spamassassin.apache.org/old/publiccorpus/ . Using these documents, we used a portion of the documents to create a decision tree data model. In order to create the decision tree data model, we used the `rpart` package for creating a decision tree and partitioning the dataset into a training and testing data. Once the decision tree data model was created, we used it to test the portion of documents that were not used in the initial creation of the decision tree. Through testing, we were able to draw final conclusions from the results.


## Creating the Dataframe

For our dataset, we took the files from https://spamassassin.apache.org/old/publiccorpus/ and converted them to a .csv file and then uploaded that to our github repository. From there, we read in the data using `readr` library and the `read_csv` function on the raw github link.

```{r}
library(tidyverse)
library(readr)
library(rpart)
library(rpart.plot)
library(caret)

url <- "https://github.com/longflin/DATA-607-Project-4/raw/refs/heads/main/emails.csv"

emails <- read_csv(url, show_col_types = FALSE)

head(emails)
```

## Email Model

Next, we created an email model.  The variables for classifying spam or ham emails include email length, number of words, numbers of exclamation marks, number of dollar signs, number of links, number of digits, and uppercase ratio.

```{r}
emails_model <- emails |>
  mutate(
    is_spam = factor(label, levels = c("ham", "spam")),
    text = replace_na(text, ""),
    text = iconv(text, to = "UTF-8", sub = ""),
    text_length = str_length(text),
    num_words = str_count(text, "\\S+"),
    num_exclamation = str_count(text, fixed("!")),
    num_dollar = str_count(text, fixed("$")),
    num_links = str_count(str_to_lower(text), "http|www"),
    num_digits = str_count(text, "\\d"),
    uppercase_ratio = if_else(
      text_length > 0,
      str_count(text, "[A-Z]") / text_length,
      0
    )
  ) |>
  select(
    is_spam,
    text_length,
    num_words,
    num_exclamation,
    num_dollar,
    num_links,
    num_digits,
    uppercase_ratio
  )
```

## Creating the Data Partition

For training and testing the model, we created a partition where 70% of the data was used for training the model and then the remaining 30% of the data was used for testing the data.

```{r}
set.seed(8585)

index <- createDataPartition(
  emails_model$is_spam,
  p = 0.7,
  list = FALSE
)

train_data <- emails_model[index, ]
test_data <- emails_model[-index, ]
```

## Decision Tree

We created a decision tree by using the 70% partitioned training data.

```{r}
spam_model <- rpart(
  is_spam ~ .,
  data = train_data,
  method = "class",
  control = rpart.control(cp = 0.01)
)

rpart.plot(
  spam_model,
  type = 5,
  extra = 101,
  under = TRUE,
  cex = 0.8,
  main = "Spam Detection Decision Tree"
)
```
## Predictions

Next we used the decision tree on the 30% partitioned testing data.

```{r}
predictions <- predict(
  spam_model,
  newdata = test_data,
  type = "class"
)

confusionMatrix(predictions, test_data$is_spam)
```

## Conclusion

The decision tree model achieved an accuracy of `0.8553` on the testing data, meaning that about 85.5% of the emails were correctly classified as either ham or spam. The confusion matrix shows that the model correctly classified 660 ham emails and 339 spam emails. However, it misclassified 79 spam emails as ham and 90 ham emails as spam. The model’s accuracy is higher than the no information rate of 0.6421, which suggests that the model performs better than simply predicting the majority class.

Balanced accuracy was 0.8455, which averages the model’s sensitivity and specificity. This is useful because the dataset has more ham emails than spam emails. The balanced accuracy shows that the model was able to predict both the the ham and the spam reasonably well. After taking into account the model's ability to distinguish between the “ham” and “spam” categories, its average performance was approximately 84.55%.

The Kappa value was 0.687 which means that there is a fairly strong agreement between the prediction and the true label.





